Global patterns of forest autotrophic carbon fluxes

### 2020-08-06 manuscript/manuscript draft.Rmd

\newpage
###Abstract
Carbon (C) fixation, allocation, and metabolism by trees set the basis for energy and material flows in forest ecosystems and define their interactions with Earth's changing climate. 
However, we lack a cohesive synthesis on how forest carbon fluxes vary globally with respect to climate and one another. 
Here, we draw upon 1319 records from the Global Forest Carbon Database (ForC), representing all major forest types and the nine most significant autotrophic carbon fluxes, to comprehensively explore how C cycling in mature, undisturbed forests varies with latitude and climate on a global scale. 
We show that, across all flux variables analyzed, C cycling decreases continuously with absolute latitude -- a finding that confirms multiple previous studies but contradicts the idea that net primary productivity ($NPP$) of temperate forests rivals that of tropical forests. 
C flux variables generally displayed similar trends across latitude and multiple climate variables, with few differences in allocation detectable at this global scale.
**Climate explained a significant proportion (24 - 71%) of variation in all C fluxes analyzed**, with temperature variables in general and mean annual temperature (MAT) in particular being the best predictors of C flux on this global scale. 
The effects of temperature were modified by moisture availability, with C flux reduced under hot and dry conditions and sometimes under very high precipitation. 
C fluxes declined with temperature seasonality, but growing season length did not improve upon MAT as a predictor.
Within the growing season, the influence of climate on C cycling was small but significant for a number of flux variables. 
These findings clarify how forest C flux varies with latitude and climate on a global scale. 
In a period of accelerating climatic change, understanding of the fundamental climatic controls on forest C cycling sets a foundation for understanding patterns of change.

@@

The latitudinal gradient in forest C flux rates, along with altitudinal gradients [@girardin_net_2010; @malhi_variation_2017], is driven primarily by climate, which is a significant driver of C fluxes across broad spatial scales [@luyssaert_co_2007; @cleveland_relationships_2011; @hursh_sensitivity_2017].
The majority of studies have focused on exploring the relationships of C fluxes to mean annual temperature (MAT) and mean annual precipitation (MAP), as the most commonly reported site-level climate variables. 
C fluxes increase strongly with $MAT$ on the global scale, but whether they saturate or potentially decrease at higher temperatures remains disputed. (**This part could use some work.**)
In the global data compilation of Luyssaert et al. [-@luyssaert_co_2007], $GPP$ increased linearly with MAT, whereas $NPP$ saturated at around 10$^\circ$C MAT. 
In contrast, Larjavaara and Muller-Landau [-@larjavaara_temperature_2012], find that increases in $GPP$ saturate at approximately 25$^\circ$C MAT, while Schuur [-@schuur_productivity_2003] finds that $NPP$ increases continuously with temperature. 
@sullivan_sensitivity_2020 found that, within the tropics, $ANPP_{stem}$ decreases at the highest maximum temperatures.
Various measures of primary productivity saturate at high levels of MAP, though the saturation points identified vary from 1500mm [@luyssaert_co_2007] up to 2,445 mm MAP [@schuur_productivity_2003]. 
Interactions between MAT and MAP are also possible; within the tropics, there is a positive interaction between MAT and MAP in shaping $ANPP$, such that high rainfall has a negative effect on productivity in cooler climates, compared to a positive effect in warmer climates [@taylor_temperature_2017]. 
Such complicated dynamics play out with belowground fluxes as well, with typically dominant soil temperature mediated by moisture and C supply [@hursh_sensitivity_2017; @xu_contribution_2016]. 
There is also evidence that C fluxes also respond to climate variables such as cloud cover [@taylor_temperature_2017], solar radiation [@beer_terrestrial_2010; @fyllas_solar_2017], and potential evapotranspiration [@kerkhoff_plant_2005]; however, these are not typically assessed in global-scale analyses of annual forest C flux.

As metrics of annual climate, MAT and MAP fail to capture variation in climate on an intra-annual scale, including temperature and precipitation seasonality and growing season length. 
Some studies have suggested that the apparently strong relationship between MAT and C fluxes is actually a consequence of the correlation between MAT and growing season length [@kerkhoff_plant_2005; @michaletz_convergence_2014; @michaletz_drivers_2018]. 
Kerkhoff et al. [-@kerkhoff_plant_2005] and Michaletz et al. [-@michaletz_convergence_2014] found no significant relationship between growing season temperature and net primary productivity (loosely defined to include a mix of **VARIABLES**) standardized to growing season length.
While this suggests that the influence of temperature may be limited to determining the length of the frost-free growing season, *analysis with clearly defined variables and standardized forest ages would be necessary to test the the veracity and generality of this hypothesis.

@@

To test whether trends in component fluxes across latitude sum to match those of larger fluxes, regression lines for smaller component fluxes were summed to generate new estimates of larger fluxes. 
Because no fluxes were significantly better predicted by a logarithmic or polynomial fit than by a linear fit, we used linear fits for all fluxes.
We then determined whether these summed predictions fell within the 95% CI for the larger flux across the entire latitudinal range.
Confidence intervals for the line of best fit for the larger flux were estimated using the 'bootMer' function, a parametric bootstrapping method for mixed models [@bates_fitting_2015]. We carried out 2000 simulations estimating the line of best fit, and took quantiles at 0.025 and 0.975 to estimate 95% CIs. **Krista please read this + confirm explanation is clear!**
This analysis was applied to the following sets of fluxes: (1) $GPP = NPP + R_{auto}$, (2) $NPP = ANPP + BNPP$, and (3) $ANPP = ANPP_{foliage} + ANPP_{stem}$. In addition, we estimated total belowground C flux (TBCF, not analyzed due to limited data) as $TBCF = BNPP + R_{root}$.

@@

All major carbon fluxes decreased with latitude (Fig. 2; Table S2). Latitude was a strong predictor for many of the carbon fluxes, particularly the larger fluxes (Table S2). Specifically, latitude explained 64% of variation in GPP (n = 243, p<0.0001), 50% in NPP (n = 161, p<0.0001) and 44% in ANPP (n = 278, p<0.0001). The C fluxes that were most poorly predicted by latitude were $BNPP_{fine.root}$ ($R^2$=0.17) and $ANPP_{stem}$ ($R^2$=0.18). The relationship with latitude was best fit by the first-order linear model, with the exception of NPP and $R_{root}$, which were best explained by the logarithmic model.

@@

In general, smaller component fluxes summed approximately to larger fluxes across the latitudinal gradient (Fig. 2). That is, modelled estimates of $GPP$, generated from the sum of $NPP$ and $R_{auto}$; $NPP$, generated from the sum of $ANPP$ and $BNPP$; and $ANPP$, generated from the sum of $ANPP_{foliage}$ and $ANPP_{stem}$, fell almost completely within the confidence intervals of the regressions of field estimates of $GPP$, $NPP$, and $ANPP$, respectively.

@@

Our analysis clarifies how forest autotrophic carbon fluxes vary with latitude and climate on a global scale, with some important implications for how forest carbon cycling relates to climate and, by extension, how it is likely to respond to climatic warming. 
Our findings show that higher temperatures with similar moisture availability result in a generalized acceleration of forest C cycling (Figs. 2-3).
This is consistent with observations of continental- to global-scale increases over time in $GPP$ [@li_mapping_2019] and $ANPP_{stem}$ [@brienen_long-term_2015; @hubau_asynchronous_2020], along with some C cycle components not considered here--tree mortality [@brienen_long-term_2015; @mcdowell_drivers_2018], soil respiration [@bond-lamberty_global_2010], and heterotrophic soil respiration [@bond-lamberty_globally_2018]. 
**DISCUSS/CITE [Wu et al. 2020]** [@wu_accelerated_2020] 
However, increasing C flux rates are by no means universal [e.g., @rutishauser_testing_2020; @hubau_asynchronous_2020], likely because other factors are at play, including changes to other aspects of climate, atmospheric pollution (CO~2~, SO~2~, NO~x~), and local disturbances.  
**(discuss new Sullivan paper that finds higher max temperatures associated to lower ANPP_stem)** [@sullivan_sensitivity_2020]
Moreover, forest ecosystem responses to climatic changes outside the temperature range to which forest communities are adapted and acclimatized will not necessarily parallel responses across geographic gradients in climate. 
Nevertheless, understanding of the fundamental climatic controls on C cycling sets a firmer foundation for understanding forest C cycle responses to accellerating climate change.

>>>>

\newpage
###Abstract
Carbon (C) fixation, allocation, and metabolism by trees set the basis for energy and material flows in forest ecosystems and define their interactions with Earth's changing climate. 
However, we lack a cohesive synthesis on how forest carbon fluxes vary globally with respect to climate and one another. 
Here, we draw upon 1319 records from the Global Forest Carbon Database (ForC), representing all major forest types and the nine most significant autotrophic carbon fluxes, to comprehensively explore how C cycling in mature, undisturbed forests varies with latitude and climate on a global scale. 
We show that, across all flux variables analyzed, C cycling decreases continuously with absolute latitude -- a finding that confirms multiple previous studies but contradicts the idea that net primary productivity ($NPP$) of temperate forests rivals that of tropical forests. 
C flux variables generally displayed similar trends across latitude and multiple climate variables, with no differences in allocation detected at this global scale.
Temperature variables in general, and mean annual temperature ($MAT$) and temperature seasonality in particular, were the best univariate predictors of C flux, explaining 19 - 71% of variation in the C fluxes analyzed. 
The effects of temperature were modified by moisture availability, with C flux reduced under hot and dry conditions and sometimes under very high precipitation. 
C fluxes increased with growing season length, but this was never the best univariate predictor.
Within the growing season, the influence of climate on C cycling was small but significant for a number of flux variables. 
These findings clarify how forest C flux varies with latitude and climate on a global scale. 
In a period of accelerating climatic change, this improved understanding of the fundamental climatic controls on forest C cycling sets a foundation for understanding patterns of change.

@@

The latitudinal gradient in forest C flux rates, along with altitudinal gradients [@girardin_net_2010; @malhi_variation_2017], is driven primarily by climate, which is a significant driver of C fluxes across broad spatial scales [@luyssaert_co_2007; @cleveland_relationships_2011; @hursh_sensitivity_2017].
However, there is little consensus as to the shapes of these relationships or the best predictor variables.
The majority of studies have focused on exploring the relationships of C fluxes to mean annual temperature (MAT) and mean annual precipitation (MAP), as the most commonly reported site-level climate variables. 
C fluxes increase strongly with $MAT$ on the global scale, but whether they saturate or potentially decrease at higher temperatures remains disputed. 
Some studies have detected no deceleration or decline in $GPP$ [@luyssaert_co_2007] or $NPP$ [@schuur_productivity_2003] with increasing MAT.
Some have suggested that relationship between C flux and temperature is exponential in nature [@allen_linking_2005] (**REFS**).
In contrast, others have found evidence of saturation or decline of C flux in the warmest climates; @luyssaert_co_2007 found $NPP$ saturating at around 10$^\circ$C MAT; @larjavaara_temperature_2012 found that increases in $GPP$ saturate at approximately 25$^\circ$C MAT, and @sullivan_sensitivity_2020 found that, within the tropics, $ANPP_{stem}$ decreases at the highest maximum temperatures.
Various measures of primary productivity saturate at high levels of MAP, though the saturation points identified vary from 1500mm [@luyssaert_co_2007] up to 2,445 mm MAP [@schuur_productivity_2003]. 
Interactions between MAT and MAP are also possible; within the tropics, there is a positive interaction between MAT and MAP in shaping $ANPP$, such that high rainfall has a negative effect on productivity in cooler climates, compared to a positive effect in warmer climates [@taylor_temperature_2017]. 
**Such complicated dynamics play out with belowground fluxes as well, with typically dominant soil temperature mediated by moisture and C supply [@hursh_sensitivity_2017; @xu_contribution_2016]. https://github.com/forc-db/Global_Productivity/issues/111**
There is also evidence that C fluxes also respond to climate variables such as temperature and precipitation seasonality (**REFS**), cloud cover [@taylor_temperature_2017], solar radiation [@beer_terrestrial_2010; @fyllas_solar_2017], and potential evapotranspiration [@kerkhoff_plant_2005]; however, these are not typically assessed in global-scale analyses of annual forest C flux.

As metrics of annual climate, MAT and MAP fail to capture variation in climate on an intra-annual scale, including temperature and precipitation seasonality and growing season length. 
Some studies have suggested that the apparently strong relationship between MAT and C fluxes is actually a consequence of the correlation between MAT and growing season length [@kerkhoff_plant_2005; @michaletz_convergence_2014; @michaletz_drivers_2018]. 
Kerkhoff et al. [-@kerkhoff_plant_2005] and Michaletz et al. [-@michaletz_convergence_2014] found no significant relationship between growing season temperature and net primary productivity (loosely defined to include a mix of **VARIABLES**) standardized to growing season length.
While this suggests that the influence of temperature may be limited to determining the length of the frost-free growing season, analysis with clearly defined variables and standardized forest ages would be necessary to test the the veracity and generality of this hypothesis.

@@

To test whether trends in component fluxes across latitude sum to match those of larger fluxes, regression lines for smaller component fluxes were summed to generate new estimates of larger fluxes. 
Because no fluxes were significantly better predicted by a logarithmic or polynomial fit than by a linear fit, we used linear fits for all fluxes.
We then determined whether these summed predictions fell within the 95% CI for the larger flux across the entire latitudinal range.
Confidence intervals for the line of best fit for the larger flux were estimated using the 'bootMer' function, a parametric bootstrapping method for mixed models [@bates_fitting_2015]. This function carried out 2000 simulations estimating the line of best fit, using quantiles at 0.025 and 0.975 to estimate 95% CIs. 
This analysis was applied to the following sets of fluxes: (1) $GPP = NPP + R_{auto}$, (2) $NPP = ANPP + BNPP$, and (3) $ANPP = ANPP_{foliage} + ANPP_{stem}$. In addition, we estimated total belowground C flux (TBCF, not analyzed due to limited data) as $TBCF = BNPP + R_{root}$.

@@

All major carbon fluxes decreased with latitude (Fig. 2; Table S2). Latitude was a strong predictor for many of the carbon fluxes, particularly the larger fluxes (Table S2). Specifically, latitude explained 64% of variation in GPP (n = 243, p<0.0001), 50% in NPP (n = 161, p<0.0001) and 44% in ANPP (n = 278, p<0.0001). The C fluxes that were most poorly predicted by latitude were $BNPP_{fine.root}$ ($R^2$=0.17) and $ANPP_{stem}$ ($R^2$=0.18). The relationship with latitude was best fit by the first-order linear model, with the exception of NPP and $R_{root}$, for which a logarithmic model was a slightly - but not significantly - better fit.

@@

In general, smaller component fluxes summed approximately to larger fluxes across the latitudinal gradient (Fig. 2). That is, modeled estimates of $GPP$, generated from the sum of $NPP$ and $R_{auto}$; $NPP$, generated from the sum of $ANPP$ and $BNPP$; and $ANPP$, generated from the sum of $ANPP_{foliage}$ and $ANPP_{stem}$, fell almost completely within the confidence intervals of the regressions of field estimates of $GPP$, $NPP$, and $ANPP$, respectively.

@@

Our analysis clarifies how forest autotrophic carbon fluxes vary with latitude and climate on a global scale, with some important implications for how forest carbon cycling relates to climate and, by extension, how it is likely to respond to climatic warming. 
Our findings show that higher temperatures with similar moisture availability result in a generalized acceleration of forest C cycling (Figs. 2-3).
This is consistent with observations of continental- to global-scale increases over time in $GPP$ [@li_mapping_2019] and $ANPP_{stem}$ [@brienen_long-term_2015; @hubau_asynchronous_2020], along with some C cycle components not considered here--tree mortality [@brienen_long-term_2015; @mcdowell_drivers_2018], soil respiration [@bond-lamberty_global_2010], and heterotrophic soil respiration [@bond-lamberty_globally_2018]. 
However, increasing C flux rates are by no means universal [e.g., @rutishauser_testing_2020; @hubau_asynchronous_2020], likely because other factors are at play, including changes to other aspects of climate, atmospheric pollution (CO~2~, SO~2~, NO~x~), and local disturbances.  
In addition, some tropical forests are
**(discuss new Sullivan paper that finds higher max temperatures associated to lower ANPP_stem)** [@sullivan_sensitivity_2020]
Moreover, forest ecosystem responses to climatic changes outside the temperature range to which forest communities are adapted and acclimatized will not necessarily parallel responses across geographic gradients in climate. 
Nevertheless, understanding of the fundamental climatic controls on C cycling sets a firmer foundation for understanding forest C cycle responses to accellerating climate change.
